---
title: "RevBayes lab scripts"
output: html_notebook
---

Most of this lab will be run in interactively in Revbayes. However, Revbayes is currently does not have sufficient simulation and plotting functions for our purposes. Thus, we will be using R in combination with Revbayes.  

Let's load up some libraries we will need later. If you have an error, install the packages and then load them.
```{r}
library(coda)
library(ape)
library(phangorn)
```


```{r}
# You may wish to set the working directory using this line, and putting your working directory in quotes.
#setwd("path/to/your/working/directory")
```

Let's simulate some linear regression data that we will then ask Revbayes to estimate.

```{r}
set.seed(5)
n <- 100
m <- rnorm(1,0,1)
b <- rnorm(1,0,1)
sigma <- rexp(1,1)
x_obs <- runif(n, -2, 2)
y_obs <- m*x_obs + b + rnorm(n, 0, sigma)

write.csv(t(x_obs), "./data/x.csv", row.names = FALSE, col.names = FALSE, )
write.csv(t(y_obs), "./data/y.csv", row.names = FALSE, col.names = FALSE)
plot(x_obs, y_obs)
print(data.frame(m,b,sigma))

```

Now go to RevBayes and analyze this data, or use the cheat below:

```{r}
system("rb revbayes_linearregression_script.rev")

```

Once you have analyzed the data in RevBayes, read it back in. 

```{r}
run1 <- read.table("./output/linear_regression1.log", header=TRUE)
head(run1)

```


```{r}
run1 <- mcmc(run1)
coln <- c("Posterior", "Likelihood", "Prior", "m", "b", "sigma")
plot(run1[,coln])
```

```{r}
plot(0,0, type="n", xlim=c(-3, 3), ylim=c(-3,3), xlab="m", ylab="b")
lines(run1[,6], run1[,5])
points(m, b, cex=2, pch=21, bg="red")
```



```{r}
run2 <- read.table("./output/linear_regression2.log", header=TRUE)
run2 <- mcmc(run2)
plot(run2[,coln])
summary(run2[,coln])
run2 <- window(run2, 500, 10000)
plot(run2[,coln])
summary(run2[,coln])
```

```{r}
par(mfrow=c(3,1))

plot(density(run2[,"m"]), main="Slope")
abline(v=m, lty=2, col="red", lwd=3)

plot(density(run2[,"b"]), main="Intercept")
abline(v=b, lty=2, col="red", lwd=3)

plot(density(run2[,"sigma"]), main="Sigma")
abline(v=sigma, lty=2, col="red", lwd=3)

```

```{r}
plot(0,0, type="n", xlim=c(-3, 3), ylim=c(-3,3), xlab="m", ylab="b")
lines(run2[,6], run2[,5])
points(m, b, cex=2, pch=21, bg="red")
```


```{r}
par(mfrow=c(2,2))
lm1 <- lm(y_obs ~ x_obs)

plot(as.numeric(run2[,"m"]), run2[,"b"], main="Slope vs. Intercept", xlab="Slope", ylab="Intercept", pch=21, col="gray50", bg="gray50", xlim=c(-2,2), ylim=c(-2,2))
points(m, b, pch=21, bg="red", cex=2)
points(lm1$coef[2], lm1$coef[1], pch="+", col="blue", cex=2)

plot(as.numeric(run2[,"m"]), run2[,"sigma"], main="Slope vs. Error", xlab="Slope", ylab="Error", pch=21, col="gray50", bg="gray50", xlim=c(-2,2), ylim=c(0,4))
points(m, sigma, pch=21, bg="red", cex=2)
points(lm1$coef[2], summary(lm1)$sigma, pch="+", col="blue", cex=2)

plot(as.numeric(run2[,"b"]), run2[,"sigma"], main="Intercept vs. Error", xlab="Intercept", ylab="Error", pch=21, col="gray50", bg="gray50", xlim=c(-2,2), ylim=c(0,4))
points(b, sigma, pch=21, bg="red", cex=2)
points(lm1$coef[1], summary(lm1)$sigma, pch="+", col="blue", cex=2)

```

# Part II. Run a phylogenetic analysis in RevBayes. Return here to visualize the results. 

```{r}
trees <- read.tree("./output/primates_cox2_GTR.trees")
tree_MAP <- read.nexus("./output/primates_cox2_GTR_MAP.tre")
tree_MAP <- root(tree_MAP, "Galeopterus_variegatus")
tree_MAP <- addConfidences(tree_MAP, trees)
tree_MAP$node.label <- round(tree_MAP$node.label,2)
plot(tree_MAP, show.node.label=TRUE, cex=0.75)
```


```{r}
par(mfrow=c(3,3), mar=c(0,0,0,0))
for(i in round(seq(1,length(trees), length.out=9), 0)){
  plot(ladderize(root(trees[[i]], outgroup = "Galeopterus_variegatus")), cex=0.5)
}
```

```{r}
trees <- root(trees, outgroup = "Galeopterus_variegatus")
for(i in 1:length(trees)){
  trees[[i]] <- phytools::force.ultrametric(trees[[i]])
}
densiTree(trees, alpha=0.2, scaleX=FALSE, col=viridis::viridis(1002))
```

# Part III. Demo of Combined Evidence Fossilized Birth-Death Process. 

```{r}
beartrees <- read.tree("./output/bears.trees")
beartree_MAP <- read.nexus("./output/bears_MAP.tre")
beartree_MAP <- addConfidences(beartree_MAP, beartrees)
beartree_MAP$node.label <- round(beartree_MAP$node.label,2)
plot(beartree_MAP, show.node.label=TRUE, cex=0.75)
```

What data and model tells us where the fossil bears go?
What data and model likely primarily informs the backbone topology of extant species?
What data informs the branch lengths? 

```{r}
par(mfrow=c(2,1), mar=c(0,0,0,0))
for(i in c(100,101)){
  plot(ladderize(beartrees[[i]]), cex=0.75)
}
```

