---
title: "Diversification & SSE Models"
output: html_notebook
---

Let's study the diversification proces using the R package `diversitree` written by Rich FitzJohn.

```{r}
library(diversitree)
library(treeplyr)
library(TreeSim)
```

# Birth-Death Models

```{r}
set.seed(1)
tree1 <- sim.bd.age(age=100, numbsim=1, lambda=0.03, mu=0)[[1]]
tree2 <- sim.bd.age(age=100, numbsim=1, lambda=0.06, mu=0)[[1]]
par(mfrow=c(1,2))
plot(tree1)
plot(tree2)
```

`Diversitree` works by first making the likelihood functions for a particular dataset. For example, for the simulated trees above, we can define the likelihood functions below using the function `make.bd`, which makes birth-death model with two parameters. What are the parameters? 

```{r}
model.tree1 <- make.bd(tree1)
model.tree2 <- make.bd(tree2)

#Test to see if functions work.
model.tree1(c(0.1,0))
model.tree2(c(0.1, 0))
```

Now we can optimize using the function `find.mle`, providing it with our starting parameter values.
```{r}
mle_tree1 <- find.mle(model.tree1, c(0.1, 0))
mle_tree2 <- find.mle(model.tree2, c(0.1, 0))
mle_tree1
mle_tree2

```

Note that we fit a pure birth tree and fit a birth-death model. We may wish to _constrain_ parameters to test particular hypotheses. For example, let's constrain extinction rate to 0 and fit a pure-birth tree. 

```{r}
pb_tree1 <- constrain(model.tree1, mu ~ 0)
pb_tree2 <- constrain(model.tree2, mu ~ 0)
find.mle(pb_tree1, c(0.1)) #Notice we only have 1 free parameter now
find.mle(pb_tree2, c(0.1))
```

We can also feed this likelihood function to a Bayesian analysis and get an estimate of the diversification rate uncertainty.
```{r}
mcmc_tree1 <- mcmc(model.tree1, c(0.1,0), w=0.1, nsteps=5000, print.every=100)
mcmc_tree2 <- mcmc(model.tree2, c(0.1,0), w=0.1, nsteps=5000, print.every=100)

```

```{r}
cols <- c("red", "green")
par(mfrow=c(3,1), mar=c(5,4,0,1))
profiles.plot(cbind("tree1_lambda"=mcmc_tree1[,'lambda'], "tree2_lambda"=mcmc_tree2[,'lambda']), col.line=cols, legend="topright")
abline(v=c(0.03, 0.06), col=cols, lwd=2, lty=2)

profiles.plot(cbind("tree1_mu"=mcmc_tree1[,'mu'], "tree2_mu"=mcmc_tree2[,'mu']), col.line=cols, legend="topright")

profiles.plot(cbind("tree1_dr"=mcmc_tree1[,'lambda']-mcmc_tree1[,'mu'], "tree2_dr"=mcmc_tree2[,'lambda'])-mcmc_tree2[,'mu'], col.line=cols, legend="topright")
abline(v=c(0.03, 0.06), col=cols, lwd=2, lty=2)
```

# Trait models
We can use the same machinery to fit character state evolution models (CTMC) models. First let's simulate some data with and without trait dependent diversification. 
```{r}
simPars <- c(0.5, 0.2, 0.05, 0.05, 0.05, 0.05)
set.seed(10)
simBisseData <- tree.bisse(simPars, max.t = 14, x0 = 0)

hist <- history.from.sim.discrete(simBisseData, 0:1)
plot(hist, simBisseData)
```


```{r}
nbModel <- make.bisse(simBisseData, simBisseData$tip.state)
p <- starting.point.bisse(simBisseData)
nbModelMLFit <- find.mle(nbModel, p)

rbind(real = simPars, estimated = round(coef(nbModelMLFit), 2))
```

```{r}
cnbModel <- constrain(nbModel, lambda1 ~ lambda0)
cnbModel <- constrain(cnbModel, mu1 ~ mu0)

cnbModelMLFit <- find.mle(cnbModel, p[c(-1, -3)])
```

```{r}
anova(nbModelMLFit, constrained = cnbModelMLFit)
```

# Challenge: How many times have geckos evolved adhesive toepads? 

Some analyses have suggested that Geckos have multiple origins of adhesive toepads (up to 11!) while other suggest the frequent loss of toepads and a single loss. Of course, a trait such as adhesive toepads may affect diversification rates as well. To adequately test this hypothesis, we need to account for the effects of diversification. Here we have data taken from Harrington and Reeder (2017; https://onlinelibrary.wiley.com/doi/full/10.1111/jeb.13004). Here I demonstrate an ancestral state reconstruction using a standard CTMC. What's the scaled likelihood of the ancestor of all geckos having toepads? 

```{r}
dat <- read.csv("./Gecko_toes.csv")
tree <- read.nexus("./Gecko_RAxML_tree.nex")
td <- make.treedata(tree, dat)
td$phy <- chronopl(td$phy, 0)##Really crappy way of making ultrametric
plot(td$phy, show.tip.label=FALSE, type="fan")
tiplabels(pch=21, bg=td[['state']])
asr <- ace(td[['state']], td$phy, type="discrete", model="ARD")
asr
nodelabels(pie=asr$lik.anc, cex=0.3, piecol = c(0,1))
```

Tasks:
1. Fit the bisse model to these data and estimate the parameters under Maximum Likelihood

```{r}

```

2. Fit a constrained model with equal speciation and extinction rates between the two states (state-independent model). Estimate parameters and fit to the data. 

```{r}

```

3. Compare the BiSSE model to the constrained model and report the best-fitting model (if equivalent, then "best model" is the model with fewest parameters)

```{r}

```


4. Test for a single origin of toepads in the geckos by constraining the best model from above (either the BiSSE model or the state-independent model) to have the relevant transition parameter set to 0. Compare this constrained model to the original model and report the best fitting model. 

```{r}

```


5. Use the best-fitting model to reconstruct ancestral states using `asr.marginal`. 

```{r}

#asr <- asr.marginal(best_model_lik, best_model_fit$par)
#plot(td$phy, type="fan", show.tip.label=FALSE)
#tiplabels(pch=21, bg=td[['state']])
#nodelabels(pie=t(asr), piecol=0:1, cex=0.3)

```


